---
title: "Computational Musicology Portfolio"
author: "Sebastiaan Dijkstra"
date: '2022-02-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(dplyr)
library(spotifyr)
library(tidyverse)
library(knitr)
library(lubridate)
library(ggplot2)
```

```{r, include=FALSE}
kanye <- get_artist_audio_features("kanye west")
# remove non-deluxe album versions to remove duplicates
kanye <- kanye[!(kanye$album_name == "Watch The Throne" | kanye$album_name == "Donda"), ]
```


### Kanye's style progression over the years.

Kanye West's albums have seen a significant shift in style over the years. Especially in his latest works, such as __Ye__ and __Jesus is King__, which are a lot slower and differ a lot from the explosive sounds of __Yeezus__ and __Watch the Throne__. To get a better feel of how much his style changed, we will look at his albums' **danceability** and **energy** values. The values come from the Spotify API and are described as follows: 

__"Danceability describes how suitable a track is for dancing based on a combination of musical elements, including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable, and 1.0 is the most danceable."__

__"Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy."__

Looking at the plot below, both the **danceability** and **energy** have dropped significantly in the last couple of years. 

```{r, include=FALSE}
style_progression <- kanye %>%
  group_by(album_name, album_release_year) %>%
  summarize(
    avgDanceability = mean(danceability),
    avgEnergy = mean(energy)
    )

plot1 <- ggplot(style_progression, aes(x=album_release_year)) +
  geom_smooth(aes(y = avgDanceability, color = "Danceability"), se=FALSE) +
  geom_smooth(aes(y = avgEnergy, color = "Energy"), se=FALSE) +
  labs(x = "Year",
       y = "Value",
       color = "Legend",
       title = "Danceability & Energy progression of Kanye West albums through the years."
       )
```

```{r plot, message=FALSE, echo=FALSE}
plot1
```

