---
title: "The Evolution of Kanye West | Computational Musicology"
author: "Sebastiaan Dijkstra | 12251267"
date: 'March 2022'
output: 
  flexdashboard::flex_dashboard:
    storyboard: true
    theme: cosmo
    self_contained: false
    navbar:
      - { title: "Github", href: "https://github.com/SebastiaanJohn/computational-musicology-portfolio", align: right, icon: fa-github}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(dplyr)
library(spotifyr)
library(tidyverse)
library(knitr)
library(lubridate)
library(ggplot2)
```

```{r, include=FALSE}
kanye <- get_artist_audio_features("kanye west")
# remove non-deluxe album versions to remove duplicates
kanye <- kanye[!(kanye$album_name == "Watch The Throne" | kanye$album_name == "Donda"), ]
```


### Introduction: The Evolution of Kanye West

I chose **Kanye West's** music because of his **progression in style** from his first studio album _The College Dropout_, to his more recent work of _Donda_ and _Jesus is King_ is remarkable. I have yet to find an artist who changes his style that much from album to album and **successfully keeps reinventing himself**. Hence I find it very interesting to base my research on him. In addition, I find it very interesting how he uses different techniques and forms of traditional music with technology and explores philosophical ideas through his lyrics, which are often thought-provoking for those who listen closely enough to hear them.

One of the most significant comparison points would be the **discrepancies between the various albums**. For instance, what attributes made each album feel so distinct but at the same time still recognize that it is a Kanye West album. I would think that the instruments and tempo are relatively different between his first and latest work. Furthermore, his lyrics have changed a lot over the decades, such as not swearing anymore, so I presume there could also be some interesting findings. 

Spotify holds all the studio albums of Kanye West; hence I don't suspect there will be any significant limitations.

**Typical Tracks:**

1. Kanye West - Heartless (808s & Heartbreak): One of the most iconic Kanye West tracks. Lots of auto tune and minimalistic songwriting. This is probably the era of Kanye tracks that most people remember.

2. Kanye West - Bound 2 (Yeezus): Very different track from the rest of Yeezus, however, still very typical Kanye West. 
  

**Atypical Tracks:**

1. Kanye West, Sunday Service Choir - Every Hour (Jesus is King): A big departure from previous work. Jesus is King is a gospel album, with the first track _Every Hour_ being in a very different style then done before.

2. Kanye West - I Thought About Killing You (ye): Very slow song that is in sharp contrast to the bass heavy tracks released earlier by Kanye.

*** 
<iframe style="border-radius:12px" src="https://open.spotify.com/embed/playlist/37i9dQZF1DZ06evO3nMr04?utm_source=generator" width="100%" height="380" frameBorder="0" allowfullscreen="" allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture"></iframe>

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/album/2Wiyo7LzdeBCsVZiRA6vVZ?utm_source=generator" width="100%" height="380" frameBorder="0" allowfullscreen="" allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture"></iframe>

### Kanye's style progression over the years.

Kanye West's albums have seen a significant shift in style over the years. Especially in his latest works, such as __Ye__ and __Jesus is King__, which are a lot slower and differ a lot from the explosive sounds of __Yeezus__ and __Watch the Throne__. To get a better feel of how much his style changed, we will look at his albums' **danceability** and **energy** values. The values come from the Spotify API and are described as follows: 

__"Danceability describes how suitable a track is for dancing based on a combination of musical elements, including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable, and 1.0 is the most danceable."__

__"Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy."__

Looking at the plot below, both the **danceability** and **energy** have dropped significantly in the last couple of years. 

```{r, include=FALSE}
style_progression <- kanye %>%
  group_by(album_name, album_release_year) %>%
  summarize(
    avgDanceability = mean(danceability),
    avgEnergy = mean(energy)
    )

plot1 <- ggplot(style_progression, aes(x=album_release_year)) +
  geom_smooth(aes(y = avgDanceability, color = "Danceability"), se=FALSE) +
  geom_smooth(aes(y = avgEnergy, color = "Energy"), se=FALSE) +
  labs(x = "Year",
       y = "Value",
       color = "Legend",
       title = "Danceability & Energy progression of Kanye West albums through the years."
       )
```

```{r plot, message=FALSE, echo=FALSE}
plot1
```

